Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-01-07 Manuscript/TEM_draft.tex
Edits to results and methods

% NC: THIS SECTION NEEDS WORK!
In this study, we focus on the effect of missing data on our ability to recover a correct topology because it is a crucial aspect of a phylogeny in many macroevolutionary studies, for example when trying to elucidate the evolutionary relationships among species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}, or for studying evolutionary transitions \citep[e.g.][]{}. %CITE-check Lloyd and Friedman or birdy stuff
Although branch length estimation is also important (namely for timing extinction and/or speciation events - e.g. \citealt{ronquista2012}), we do not consider branch lengths in this study. This is partially due to difficulties of estimating the effect of missing data simultaneously on branch length and topology, % TG: or "This is partially due to difficulties with simulating branch lengths and topology simultaneously"
but also because previous studies have already empirically assessed the effect of the Total Evidence method on branch length variation but with a fixed topology approach \citep{ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}.
Thus understanding the sensitivity of topology to missing data is important for assessing the accuracy of tree estimation in the Total Evidence framework. To our knowledge, this question has never been formally assessed.

@@

\subsubsection{Robinson-Foulds distance}
The Robinson-Foulds distance \citep{RF1981}, or "path difference", measures the number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees (\citealt{RF1981}; see \hyperref[SupplementaryMaterial]{Supplementary Material S2} for calculation details). This metric is bounded between one, when the two trees are identical, and $n-2$ (for two trees with $n$ taxa) when there is no shared clade in the two trees. This metric is sensitive to minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity. We normalised this metric using the Normalized Tree Similarity index \citep{Bogdanowicz2012}. This index allows us to compare the metric for any number of taxa by scaling it with the expected value of the metric when comparing two random trees (see \hyperref[SupplementaryMaterial]{Supplementary Material S2} for the calculation details). Therefore, when the Normalised Robinson-Foulds distance between two trees is equal to one, the trees are identical; if the Normalised Robinson-Foulds distance is equal to zero, the trees are no more different than expected by chance; finally if the Normalised Robinson-Foulds distance is lower than zero, the trees are more different than expected by chance.

\subsubsection{Triplets distance}
The Triplets distance \citep{dobson1975triplets} measures the number of sub-trees made up of three taxa that differ between two trees (\citealt{critchlowthe1996}; see \hyperref[SupplementaryMaterial]{Supplementary Material S2} for calculation details). This metric measures the position of each taxon and clade in relation to its closest neighbours. It is bounded between 0 when the two trees are identical and $\binom{n}{4}$ (for two trees with $n$ taxa) when there is no shared taxa/clade position in the two trees. Therefore this metric sensitive to the conservation of wildcard taxa. We normalised this metric in the same way as we did for the Robinson-Foulds distance resulting in the Normalised Triplets distance.

%\subsubsection{Normalized Tree Similarity}
%The Normalized Tree Similarity index, $NTS_m$ (\citealp{Bogdanowicz2012}) allows us to compare the two metrics above for any number of taxa, by scaling the value of the metric $m$ (either Robinson-Foulds or Triplets distance in our study) to the expected value of the metric $m$ when comparing two random trees (see \hyperref[SupplementaryMaterial]{Supplementary Material S2}). When $NTS_m$=1, the two trees are identical, when $NTS_m$=0 the trees are no more different than expected when comparing two random trees, and when $NTS_m$$<$0, the difference between the two trees is greater than when comparing two random trees. In our study we used the $NTS_m$ index as a proxy for topological similarity: a high score of this index (i.e. towards one) means that the topology is highly conserved between the two trees; conversely, a low score of this index (i.e. towards 0) means that the topological difference between the two trees is as great as expected when comparing two random trees. Note that the two tree comparison metrics represent different aspects of topological similarity and do not perform equally well (see \hyperref[metrics_discussion]{discussion below} and \citealt{kuhnerpractical2014}).

@@

\subsubsection{Summarizing the comparisons}
Finally after doing all theses %stupid
50 simulations and comparisons, we tested the effects of our missing data parameters ($M_{L}$, $M_{F}$ and $M_{C}$) on the ability to recover the correct topology in a Total Evidence framework. Additionally, we also assessed the effect of the different methods (i.e. Maximum Likelihood, Bayesian consensus, Bootstraps and Bayesian posterior trees distribution).% and the joint effect of the our missing data parameters and the methods.
% TG: the following part might be useless (complete repetition of above)
 To assess the ability to recover the "correct" tree topology on our 50 simulations, we used the scores for both normalised metrics (Robinson-Foulds and Triplets distance). We looked at the following effects:
\begin{enumerate}
\item{The three missing data parameters ($M_{L}$, $M_{F}$ and $M_{C}$);}
\item{The four "methods" (Maximum Likelihood, Bayesian consensus, Bootstraps and Bayesian posterior trees distribution);}
%\item{The joint effect of the missing data parameters and the methods.} % actually this is implied in both already.
\end{enumerate}
% TG: end of the repetition

%Introducing Bhattacharyya Coefficient
To evaluate these effects we measured the similarities among the different normalised metrics scores (Robinson-Foulds and Triplets distance) for all the analysis above using the Bhattacharyya Coefficient \citep{Bhattacharyya}. The Bhattacharyya Coefficient is the probability of overlap between two distributions, ranging from 0 to 1 (\citealt{Bhattacharyya} - see \hyperref[SupplementaryMaterial]{Supplementary Material Section 2} for calculation details).

%Explaining the parameter pairwise comparisons (ANOVA equivalent=within)
For assessing the effect of our three missing data parameters ($M_{L}$, $M_{F}$ and $M_{C}$), we calculated the Bhattacharyya Coefficient between each pairwise combination for each parameter state (0\%, 10\%, 25\%, 50\%, 75\%). This resulted in 7875 pairwise distributions pairwise comparisons (triangular $3^5$$\times$$3^5$ matrix). We represented the results in a triangular matrix with the values of each pairwise Bhattacharyya Coefficient coloured according to their values (in green when the distribution overlap completely (Bhattacharyya Coefficient=1) and in red when they don't (Bhattacharyya Coefficient=0) - Fig.~\ref{Fig_Results-paircomp_within}). We repeated this procedure for all the "methods" (Maximum Likelihood, Bayesian consensus, Bootstraps and Bayesian posterior trees distribution).

@@

%Questionish §
We use the Bhattacharyya Coefficient in a similar way as a two-sided t-test to detect significant differences between the distributions of the normalised distance between the "best" tree and the "missing data" trees (for both metrics).
%We also take into account the effect of the methods used. REPHRASE
This way we could investigate the effect of each our three missing data parameters and their interaction on the ability to recover the "correct" topology.

@@

%Part 2
\subsection{Effect of missing living taxa ($M_{L}$)} %Or remove? Also below then
As the amount of missing living taxa increases, the ability to recover the correct topology decreases (Fig.~\ref{Fig_Results-permeth_perparam}). However, the decrease of the Normalised Robinson-Fould distance plateaus at minimum of 0.69 for Bayesian consensus tree (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam} - see also \hyperref[SupplementaryMaterial]{Supplementary Material Section 3} for the exact values) and there is no substantial difference in topological recovery for 50\% or 75\% of missing living taxa (upper triangle in Fig.~\ref{Fig_Results-paircomp_within}-A). Therefore, clades seems to be more conserved in the Bayesian consensus tree more than any other method (see discussion).  %add discussion link
On the other hand, the decrease in Normalised Triplets distance is less contrasted with a greater overlap of Normalised Triplets distance with the Maximum Likelihood tree (Fig.~\ref{Fig_Results-permeth_perparam}) - mean Bhattacharrya Coefficient=0.10 for the Normalised Robinson-Foulds distance and 0.61 for the Normalised Triplets distance - see table ~\ref{Tab_Results-Difference_methods}).

>>>>

In this study, we focus on the effect of missing data on our ability to recover a correct topology because it is a crucial aspect of a phylogeny in many macroevolutionary studies, for example when trying to elucidate the evolutionary relationships among species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}, or for studying evolutionary transitions \citep[e.g.][]{}. %CITE-check Lloyd and Friedman or birdy stuff
Although branch length estimation is also important (namely for timing extinction and/or speciation events - e.g. \citealt{ronquista2012}), we do not consider branch lengths in this study. This is partially due to difficulties with simulating branch lengths and topology simultaneously,
but also because previous studies have already empirically assessed the effect of the Total Evidence method on branch length variation but with a fixed topology approach \citep{ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}.
Thus understanding the sensitivity of topology to missing data is important for assessing the accuracy of tree estimation in the Total Evidence framework. To our knowledge, this question has never been formally assessed.

@@

\subsubsection{Robinson-Foulds distance}
The Robinson-Foulds distance \citep{RF1981}, or "path difference", measures the number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees (\citealt{RF1981}; see \hyperref[SupplementaryMaterial]{Supplementary Material S2} for calculation details). This metric is bounded between one, when the two trees are identical, and $n-2$ (for two trees with $n$ taxa) when there is no shared clade in the two trees. This metric is sensitive to minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity. We normalised this metric by % NC: DETAILS 
\citep{Bogdanowicz2012}. This allows us to compare the metric for any number of taxa by scaling it with the expected value of the metric when comparing two random trees (see \hyperref[SupplementaryMaterial]{Supplementary Material S2} for the calculation details). Therefore, when the Normalised Robinson-Foulds distance between two trees is equal to one, the trees are identical; if the distance is equal to zero, the trees are no more different than expected by chance; finally if the distance is less than zero, the trees are more different than expected by chance.

\subsubsection{Triplets distance}
The Triplets distance \citep{dobson1975triplets} measures the number of sub-trees made up of three taxa that differ between two trees (\citealt{critchlowthe1996}; see \hyperref[SupplementaryMaterial]{Supplementary Material S2} for calculation details). This metric measures the position of each taxon and clade in relation to its closest neighbours. It is bounded between 0 when the two trees are identical and $\binom{n}{4}$ (for two trees with $n$ taxa) when there is no shared taxa/clade position in the two trees. Therefore this metric sensitive to the conservation of wildcard taxa. We normalised this metric in the same way as for the Robinson-Foulds distance resulting in the Normalised Triplets distance.

@@

\subsubsection{Summarizing the comparisons}
Finally, we tested the effects of our missing data parameters ($M_{L}$, $M_{F}$, $M_{C}$ and their interactions) on our ability to recover the correct topology in a Total Evidence framework. We also assessed the effect of our missing data parameters jointly with the effects of different tree inference and uncertainty methods (i.e. Maximum Likelihood, Bayesian consensus, Maximum Likelihood bootstrap trees and Bayesian posterior tree distribution).

%Introducing Bhattacharyya Coefficient
To evaluate these effects we measured similarities among different metrics scores (Normalised Robinson-Foulds and Normalised Triplets distance) for all the analyses above using the Bhattacharyya Coefficient \citep{Bhattacharyya}. The Bhattacharyya Coefficient is the probability of overlap between two distributions (\citealt{Bhattacharyya} - see \hyperref[SupplementaryMaterial]{Supplementary Material Section 2} for calculation details).

%Explaining the parameter pairwise comparisons (ANOVA equivalent=within)
For assessing the effect of our missing data parameters, we calculated the Bhattacharyya Coefficient between the Normalised Robinson-Foulds and Normalised Triplets distance distributions, for each pairwise combination of missing data parameters ($M_{L}$, $M_{F}$, $M_{C}$) and parameter states (0\%, 10\%, 25\%, 50\%, 75\%). For example, $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%; $M_{L}$ = 10\%, $M_{F}$ = 0\%, $M_{C}$ = 0\% etc. This resulted in 7875 pairwise comparisons (a triangular matrix with $3^5$$\times$$3^5$ cells). We represented the results in a triangular matrix with the values of each pairwise Bhattacharyya Coefficient coloured according to their values (in green when the distribution overlap completely (Bhattacharyya Coefficient = 1) and in red when they don't (Bhattacharyya Coefficient = 0) - Fig.~\ref{Fig_Results-paircomp_within}). We repeated this procedure for all the "methods" (Maximum Likelihood, Bayesian consensus, Bootstraps and Bayesian posterior trees distribution).

@@

%Questionish §
We use the Bhattacharyya Coefficient in a similar way as a two-sided t-test to detect significant differences between the distributions of the normalised distance between the "best" tree and the "missing data" trees (for both metrics).
%We also take into account the effect of the methods used. REPHRASE
%This way we could investigate the effect of each our three missing data parameters and their interaction on the ability to recover the "correct" topology.

@@

%Part 2
\subsection{Individual effects of missing data parameters} 
As the amount of missing data increases across all three parameters, the ability to recover the correct topology decreases (Fig.~\ref{Fig_Results-permeth_perparam}). However, the decrease of the Normalised Robinson-Fould distance plateaus at minimum of 0.69 for Bayesian consensus tree (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam} - see also \hyperref[SupplementaryMaterial]{Supplementary Material Section 3} for the table of values) and there is no substantial difference in topological recovery for 50\% or 75\% of missing living taxa (upper triangle in Fig.~\ref{Fig_Results-paircomp_within}-A). Therefore, clades seems to be more conserved in the Bayesian consensus tree more than any other method (see discussion).  %add discussion link
On the other hand, the decrease in Normalised Triplets distance is less contrasted with a greater overlap of Normalised Triplets distance with the Maximum Likelihood tree (Fig.~\ref{Fig_Results-permeth_perparam}) - mean Bhattacharrya Coefficient=0.10 for the Normalised Robinson-Foulds distance and 0.61 for the Normalised Triplets distance - see table ~\ref{Tab_Results-Difference_methods}).
